Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2014-12-10 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Updated figures in wgbs_bioinformatics_analysis

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption{texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveal the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

Generally, when referring to m-tuples I implicitly mean $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple. The default option of `methtuple` is to produce m-tuples with $NIL = 0$ unless the `--all-combination` flag is set[^nil0].

@@

To illustrate several of the above-mentioned concepts, suppose we sequence a region of the genome containing five methylation loci with three paired-end reads (`A`, `B` and `C`):

__TODO: Convert the following code chunks to figures in Illustrator__

```
ref: 1    2   3 4 5
A_1: |----->
A_2:         <------|
B_1:
B_2: |----->   <----|
C_1:    |----->
C_2:      <------|
```

If we are interested in 1-tuples, then we would obtain the following from each read by running `methtuple`:

```
A: {1}, {2}, {3}, {4}, {5}
B: {1}, {2}, {4}, {5}
C: {2}, {3}, {4}
```

The result is identical regardless of whether the `--all-combinations` flag is set.

If we are interested in 3-tuples, then we would obtain the following from each read by running `methtuple` in its default mode:

```
A: {1, 2, 3}, {2, 3, 4}, {3, 4, 5}
B: {1, 2, 4}, {2, 4, 5}
C: {2, 3, 4}
```

Things to note:

- Read-pair `A` sequences all three (= 5 - 3 + 1) adjacent 3-tuples
- Read-pair `B` sequences none of the adjacent 3-tuples but does 'erroneously' construct two 3-tuples with $NIL > 0$. This happens because m-tuples are created independently from each read-pair; effectively, read-pair `B` is unaware of methylation locus `3`. Depending on the downstream analysis, you may want to _post-hoc_ filter out these m-tuples with $NIL > 0$.
- The twice-sequenced methylation loci, `2` and `3`, in read-pair `C` are not double counted.

However, if we were to run `methtuple` with `--all-combinations` then we would obtain:

```
A: {1, 2, 3}, {2, 3, 4}, {3, 4, 5}, {1, 2, 4}, {1, 2, 5}, {1, 3, 4}, {1, 3, 5}, {1, 4, 5}, {2, 3, 5}, {2, 4, 5}
B: {1, 2, 4}, {2, 4, 5}, {1, 2, 5}, {1, 4, 5}
C: {2, 3, 4}
```

The number of CpGs per read for the Lister data is shown in Figure \ref{fig:Lister_cpgs_per_read}. Longer reads, and paired-end reads, contain more methylation loci and so are more informative for analyses using m-tuples. This can be seen by comparing, for example, the _ADS_ and _HSF1_ samples. Samples sequenced more deeply will have more reads per m-tuple, although this can't be seen in these plots since they are normalised by sequencing depth.

```{r Lister_cpgs_per_read, dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data.", out.width = "0.8\\linewidth"}
library(dplyr)
library(ggplot2)
library(scales)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
dataset <- "Lister"
hist <- lapply(paste0("../data/observable_m-tuples/", dataset, "/",
                      list.files(paste0("../data/observable_m-tuples/",
                                        dataset), "\\.CG\\_per\\_read\\.hist")),
               read.table, header = TRUE)
sn <- gsub("\\.CG\\_per\\_read\\.hist", "",
           list.files(paste0("../data/observable_m-tuples/",
                             dataset), "\\.CG\\_per\\_read\\.hist"))
sn <- gsub("DM\\_", "", sn)
hist_df <- bind_rows(hist) %>%
  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
  group_by(Sample) %>%
  mutate(Proportion = count / sum(as.numeric(count)))
hist_df %>%
  ggplot(aes(x = n, y = Proportion * 100)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Sample, ncol = 5, nrow = 4) +
  coord_cartesian(xlim = c(-1, 8.5)) +
  scale_y_continuous(label = comma) +
  xlab("Number of CpGs per read\n(Axis truncated at 8)") +
  ylab("Percentage of total reads") +
  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
  thesis_theme
```

@@

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
\texttt{methclone} & \citet{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Try installing. Suspect you can't get actual methylation calls.} \\
\texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
\texttt{DMEAS}     & \citet{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only. Perl code only available as PDF file.             \\ \bottomrule
\end{tabulary}
\caption{Other software for methylation calling at m-tuples.}
\label{tab:other_m-tuple_methylation_callers}
\end{table}

@@

The output format of `methtuple` is tab-delimited plain text, optionally compressed with `gzip` or `bzip2`. An example of the output is shown below for 3-tuples:

```{sh}
chr     strand  pos1    pos2    pos3    MMM     MMU     MUM     MUU     UMM     UMU     UUM     UUU
chr1    +       469     471     484     0       0       4       1       1       0       0       0
chr1    +       471     484     489     1       0       0       0       2       2       1       0

```

>>>>

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption{texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveal the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

Generally, when referring to m-tuples I implicitly mean $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple. The default option of `methtuple` is to produce m-tuples with $NIL = 0$ unless the `--all-combinations` flag is set[^nil0].

@@

To illustrate several of the above-mentioned concepts, suppose we sequence a region of the genome containing five methylation loci with three paired-end reads (`A`, `B` and `C`), shown in Figure \ref{fig:methtuple_example1}.

\begin{figure}[h!]
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}.}
\label{fig:methtuple_example1}
\end{figure}

If we are interested in 1-tuples, Figure \ref{fig:methtuple_example2} shows what we would obtain from each read by running `methtuple`. The result is identical regardless of whether the `--all-combinations` flag is set.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
\caption{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example2}
\end{figure}

If we are interested in 3-tuples, Figure \ref{fig:methtuple_example3} shows what we would obtain from each read by running `methtuple` in its default mode. A few things to note:

- Read-pair `A` sequences all three (= 5 - 3 + 1) adjacent 3-tuples
- Read-pair `B` sequences none of the adjacent 3-tuples but does 'erroneously' construct two 3-tuples from pairs of non-adjacent loci. This happens because m-tuples are created independently from each read-pair; effectively, read-pair `B` is unaware of methylation locus `3`. Depending on the downstream analysis, you may want to _post-hoc_ filter out these m-tuples with non-adjacent loci.
- The twice-sequenced methylation loci in read-pair `C`, `2` and `3`, are not double counted.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example3}
\end{figure}

Finally, Figure \ref{fig:methtuple_example4} shows the output if we were to analyse 3-tuples but with the `--all-combinations` flag set.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
\label{fig:methtuple_example4}
\end{figure}

With current sequencing technology we are limited to extracting m-tuples that span no more than $200-250$ bp. This obviously affects the size of m-tuples that we can study. Figure \ref{fig:Lister_CpGs_per_read} shows the number of CpGs per read for the Lister dataset (see Chapter \ref{chap:dataset} for a description of the Lister dataset). Longer reads, and paired-end reads, contain more methylation loci and so are more informative for analyses using m-tuples. This can be seen by comparing, for example, the _ADS_ and _HSF1_ samples. Samples sequenced more deeply will have more reads per m-tuple, although this can't be seen in these plots since they are normalised by sequencing depth.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
\caption{Number of CpGs per read for the Lister dataset.}
\label{fig:Lister_CpGs_per_read}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
\texttt{methclone} & \citet{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Suspect you can't get actual methylation calls.} \\
\texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
\texttt{DMEAS}     & \citet{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only. Perl code only available as PDF file.             \\ \bottomrule
\end{tabulary}
\caption{Other software for methylation calling at m-tuples.}
\label{tab:other_m-tuple_methylation_callers}
\end{table}

@@

The output format of `methtuple` is tab-delimited plain text, optionally compressed with `gzip` or `bzip2`. Figure \ref{fig:3-tuples} shows an example of the output for 3-tuples.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
\caption{Example output of \texttt{methtuple} for 3-tuples.}
\label{fig:3-tuples}
\end{figure}
